---
title: "cm007 Exercises: Practice with `dplyr`"
output: 
  html_document:
    keep_md: true
    theme: paper
---


<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tsibble))
```


# Following along with lecture notes

##### summarize is almost never used on it's own, usually used in tandem with the group_by function.

##### Can't group_by with a dataframe, unique to tibbles
Can group by new variables, kind of like mutate

1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(smallLifeExp = lifeExp < 60) %>% 
  select(country, year, smallLifeExp)
```
Want to compute the mean and standard deviation for each year for every continent?
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```



dplyr has a bunch of convenience functions that help us write code more eloquently. We could use group_by() and summarize() with length() to find the number of entries each country has:
```{r}
gapminder %>% 
  group_by(country) %>% 
  transmute(n = length(country))
```


Or, we can use dplyr::n() to count the number of rows in each group:
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(n = n())
```

Or better yet, just use dplyr::count():
```{r}
gapminder %>% 
  count(country)
```

Want to get the increase in GDP per capita for each country?
# Arranged just to make sure years are in order...

```{r}
gap_inc <- gapminder %>% 
  arrange(year) %>%  
  group_by(country) %>%
  mutate(gdpPercap_inc = gdpPercap - lag(gdpPercap))
DT::datatable(gap_inc)
```


Drop the NAs with another convenience function, this time supplied by the tidyr package (another tidyverse package that we’ll see soon):
```{r}
gap_inc %>% 
  tidyr::drop_na()
```



# Class excercises 
# This worksheet contains exercises aimed for practice with `dplyr`. 

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(min_life = min(lifeExp),
            country = country[lifeExp == min_life]) %>% 
  arrange(min_life)
```

2. Calculate the growth in population since the first year on record _for each country_ by rearranging the following lines, and filling in the `FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 


```{r}
gapminder %>%
  group_by(country) %>%
  arrange(year) %>% 
  mutate(rel_growth = pop - first(pop)) %>% 
DT::datatable() 
```


3. *Determine the country that experienced the sharpest 5-year drop in life expectancy*, in each continent, sorted by the drop, by rearranging the following lines of code. Ensure there are no `NA`'s. Instead of using `lag()`, use the convenience function provided by the `tsibble` package, `tsibble::difference()`:

```

 
 

```

```{r}
gapminder %>%
  group_by(country) %>%
  arrange(year) %>% 
  mutate(inc_life_exp = difference(lifeExp)) %>% 
  drop_na() %>% 
  ungroup() %>%  ## because actually looking at continent but first phase was calc lifeExp in each country
  group_by(continent) %>%
  filter(inc_life_exp == min(inc_life_exp)) %>%  
  arrange(inc_life_exp) %>%
  knitr::kable()
  
```


